---
title: "STA206 Project"
author: "Xinyi Cen"
date: "2024-11-19"
output: pdf_document
---

# Import data

```{r}
# Import the data from R library
# all variable columns in lowercase
library(bsamGP)
plasma

# Import the data from txt file
# all varaible columns in uppercase
# plasma<-read.table("/Users/centracy/Desktop/plasma.txt")
# colnames(plasma)<-plasma[1,]
# plasma<-plasma[-1,]
# # Cover numerical variables as numeric, categorical variables as factor
# num<-c('AGE','QUETELET','CALORIES','FAT','FIBER',
#        'ALCOHOL','CHOLESTEROL','BETADIET','RETDIET','BETAPLASMA','RETPLASMA')
# cate<-c('SEX','SMOKSTAT','VITUSE')
# plasma[num]<-lapply(plasma[num],as.numeric)
# plasma[cate]<-lapply(plasma[cate],as.factor)
# plasma
```

```{r}
#par()
library(ggplot2)
library(patchwork)

p1<-ggplot(plasma,aes(x=factor(sex),y=retplasma))+
  geom_boxplot() +
  labs(title = paste("Boxplot of Sex vs Retplasma"), x = "Sex", y = "Plasma")
p2<-ggplot(plasma,aes(x=factor(smoke),y=retplasma))+
  geom_boxplot() +
  labs(title = paste("Boxplot of Smoke vs Retplasma"), x = "Smoke", y = "Plasma")
p3<-ggplot(plasma,aes(x=factor(vituse),y=retplasma))+
  geom_boxplot() +
  labs(title = paste("Boxplot of Vituse vs Retplasma"), x = "Vituse", y = "Plasma")
p1+p2+p3

p4<-ggplot(plasma,aes(x=factor(sex),y=betaplasma))+
  geom_boxplot() +
  labs(title = paste("Boxplot of Sex vs betaplasma"), x = "Sex", y = "Plasma")
p5<-ggplot(plasma,aes(x=factor(smoke),y=betaplasma))+
  geom_boxplot() +
  labs(title = paste("Boxplot of Smoke vs Retplasma"), x = "Smoke", y = "Plasma")
p6<-ggplot(plasma,aes(x=factor(vituse),y=betaplasma))+
  geom_boxplot() +
  labs(title = paste("Boxplot of Vituse vs betaplasma"), x = "Vituse", y = "Plasma")
p4+p5+p6
```

# Check the outliers
```{r}
# Consider exclude those cases to improve the prediction
# Check the extreme outliers from the betaplasma variable which might be an input error
plasma[which(plasma$betaplasma>1380),] # case row 40 extreme
# Check the extreme outliers from the retplasma variable which might be an input error
plasma[which(plasma$retplasma>1500),] # case row 20 extreme
```

# Model for personal characteristics vs betaplasma & retplasma
- How do personal characteristics such as age, sex, and smoking status influence plasma concentrations of beta-carotene and retinol?

**linear regression**
```{r}
# Model fit with first order
ret<-lm(retplasma~sex+age+smoke+bmi,data=plasma)
beta<-lm(betaplasma~sex+age+smoke+bmi,data=plasma)

# Assumption check
par(mfrow = c(2, 3)) # Set the plots as 2x3 layout

# Residuals VS fitted
plot(x=ret$fitted.values,y=residuals(ret),main = "Residuals VS Fitted value for retinol",
     xlab = "Fitted value",ylab = "Residuals")
abline(h=0,col='red')

# QQ plot
qqnorm(residuals(ret),main='Normal QQ-plot for retinol')
qqline(residuals(ret),col='red')

# Residual Box plot
boxplot(residuals(ret),main='Residual Box Plot for retinol')

# Residuals VS fitted
plot(x=beta$fitted.values,y=residuals(beta),main = "Residuals VS Fitted value for beta-carotene",
     xlab = "Fitted value",ylab = "Residuals")
abline(h=0,col='red')

# QQ plot
qqnorm(residuals(beta),main='Normal QQ-plot for beta-carotene')
qqline(residuals(beta),col='red')

# Residual Box plot
boxplot(residuals(beta),main='Residual Box Plot for beta-carotene')


summary(ret)
summary(beta)
anova(ret)
anova(beta)
```

**Interaction term linear regression**
```{r}
# Adding interaction terms in the linaer regression model
ret1<-lm(retplasma~sex+age+smoke+sex:age+sex:smoke+smoke:age+sex:bmi+smoke:bmi+age:bmi,data=plasma)
beta1<-lm(betaplasma~sex+age+smoke+sex:age+sex:smoke+smoke:age+sex:bmi+smoke:bmi+age:bmi,data=plasma)

# Assumption check
par(mfrow = c(2, 3)) # Set the plots as 1x3 layout

# Residuals VS fitted
plot(x=ret1$fitted.values,y=residuals(ret1),main = "Residuals VS Fitted value for retinol",
     xlab = "Fitted value",ylab = "Residuals")
abline(h=0,col='red')

# QQ plot
qqnorm(residuals(ret1),main='Normal QQ-plot for retinol')
qqline(residuals(ret1),col='red')

# Residual Box plot
boxplot(residuals(ret1),main='Residual Box Plot for retinol')

# Residuals VS fitted
plot(x=beta1$fitted.values,y=residuals(beta1),main = "Residuals VS Fitted value for beta-carotene",
     xlab = "Fitted value",ylab = "Residuals")
abline(h=0,col='red')

# QQ plot
qqnorm(residuals(beta1),main='Normal QQ-plot for beta-carotene')
qqline(residuals(beta1),col='red')

# Residual Box plot
boxplot(residuals(beta1),main='Residual Box Plot for beta-carotene ')

summary(ret1)
summary(beta1)
anova(ret1)
anova(beta1)
```

**Log regression model**
```{r}
# Preparation for the log regression model selected by ANOVA test
anova(ret,ret1)
anova(beta,beta1)
# Adding interaction terms in the linear regression model
ret2<-lm(log(retplasma)~sex+age+smoke+bmi,data=plasma)
beta2<-lm(log(betaplasma)~sex+age+smoke+bmi,data=plasma)

# Assumption check
par(mfrow = c(2, 3)) # Set the plots as 1x3 layout

# Residuals VS fitted
plot(x=ret2$fitted.values,y=residuals(ret2),main = "Residuals VS Fitted value for retinol",
     xlab = "Fitted value",ylab = "Residuals")
abline(h=0,col='red')

# QQ plot
qqnorm(residuals(ret2),main='Normal QQ-plot for retinol')
qqline(residuals(ret2),col='red')

# Residual Box plot
boxplot(residuals(ret2),main='Residual Box Plot for retinol')

# Residuals VS fitted
plot(x=beta2$fitted.values,y=residuals(beta2),main = "Residuals VS Fitted value for beta-carotene",
     xlab = "Fitted value",ylab = "Residuals")
abline(h=0,col='red')

# QQ plot
qqnorm(residuals(beta2),main='Normal QQ-plot for beta-carotene')
qqline(residuals(beta2),col='red')

# Residual Box plot
boxplot(residuals(beta2),main='Residual Box Plot for beta-carotene ')

summary(ret2)
summary(beta2)
anova(ret2)
anova(beta2)
```

**Model Selection**
```{r}
# AIC
aic_r<-c(AIC(ret),AIC(ret1),AIC(ret2))
aic_b<-c(AIC(beta),AIC(beta1),AIC(beta2))

#BIC
bic_r<-c(BIC(ret),BIC(ret1),BIC(ret2))
bic_b<-c(BIC(beta),BIC(beta1),BIC(beta2))


comp1<-data.frame(Model = c("Simple Linear", "Interaction Terms", "Log-Transformed"),
  AIC_ret = aic_r,
  BIC_ret = bic_r,
  AIC_beta=aic_b,
  BIC_beta=bic_b)
comp1


# Outliers concern
library(car)
par(mfrow=c(2,2))
term<-c('sex','age','smoke','bmi')
for (i in term) {
    leveragePlot(ret2, term.name = i,id = list(n = 2, col = "red"))
}

par(mfrow=c(2,2))
for (i in term) {
    leveragePlot(beta2, term.name = i,id = list(n = 2, col = "red"))
}


```

# Dietary intake with betaplasma (included intake of dietary betaplasma) and retplasma (included intake of retplasma)
- Is there a relationship between dietary intake of beta-carotene? If so, what is the relationship between dietary intake of beta-carotene and plasma concentrations of beta-carotene?
- Is there a linear relationship between total caloric intake and plasma levels of beta-carotene (betaplasma)?

```{r}
rdt<-lm(retplasma~vituse+calories+fat+fiber+alcohol+cholesterol+retdiet+betadiet,data=plasma)
bdt<-lm(betaplasma~vituse+calories+fat+fiber+alcohol+cholesterol+retdiet+betadiet,data=plasma)

# Assumption check
par(mfrow = c(2, 3)) # Set the plots as 1x3 layout

# Residuals VS fitted
plot(x=rdt$fitted.values,y=residuals(rdt),main = "Residuals VS Fitted value for retinol",
     xlab = "Fitted value",ylab = "Residuals")
abline(h=0,col='red')

# QQ plot
qqnorm(residuals(rdt),main='Normal QQ-plot for retinol')
qqline(residuals(rdt),col='red')

# Residual Box plot
boxplot(residuals(rdt),main='Residual Box Plot for retinol')

# Residuals VS fitted
plot(x=bdt$fitted.values,y=residuals(bdt),main = "Residuals VS Fitted value for beta-carotene",
     xlab = "Fitted value",ylab = "Residuals")
abline(h=0,col='red')

# QQ plot
qqnorm(residuals(bdt),main='Normal QQ-plot for beta-carotene')
qqline(residuals(bdt),col='red')

# Residual Box plot
boxplot(residuals(bdt),main='Residual Box Plot for beta-carotene')

summary(rdt)
anova(rdt)
summary(bdt)
anova(bdt)
```

**Interaction term model**
```{r}
# Preparation for log regression model selected terms by using stepwise AIC
library(MASS)
stepAIC(rdt,scope = 
          list(
  lower = ~ vituse + calories + fat + fiber + alcohol + cholesterol + retdiet + betadiet,
  upper = ~ (vituse + calories + fat + fiber + alcohol + cholesterol + retdiet + betadiet)^2),
        direction="both",k=2,trace=FALSE)$anova

stepAIC(bdt,scope =list(
  lower = ~ vituse + calories + fat + fiber + alcohol + cholesterol + retdiet + betadiet,
  upper = ~ (vituse + calories + fat + fiber + alcohol + cholesterol + retdiet + betadiet)^2),
        direction="both",k=2,trace=FALSE)$anova

rdt1<-lm(retplasma~vituse+calories+fat+fiber+alcohol+cholesterol+retdiet+betadiet
         +calories:alcohol+fat:fiber+vituse:betadiet,data=plasma)
bdt1<-lm(betaplasma~vituse+calories+fat+fiber+alcohol+cholesterol+retdiet+betadiet+
           vituse:fiber+fiber:betadiet+fat:betadiet+fat:alcohol+calories:alcohol+vituse:alcohol,
         data=plasma)

# Assumption check
par(mfrow = c(2, 3)) # Set the plots as 1x3 layout

# Residuals VS fitted
plot(x=rdt1$fitted.values,y=residuals(rdt1),main = "Residuals VS Fitted value for retinol",
     xlab = "Fitted value",ylab = "Residuals")
abline(h=0,col='red')

# QQ plot
qqnorm(residuals(rdt1),main='Normal QQ-plot for retinol')
qqline(residuals(rdt1),col='red')

# Residual Box plot
boxplot(residuals(rdt1),main='Residual Box Plot for retinol')

# Residuals VS fitted
plot(x=bdt1$fitted.values,y=residuals(bdt1),main = "Residuals VS Fitted value for beta-carotene",
     xlab = "Fitted value",ylab = "Residuals")
abline(h=0,col='red')

# QQ plot
qqnorm(residuals(bdt1),main='Normal QQ-plot for beta-carotene')
qqline(residuals(bdt1),col='red')

# Residual Box plot
boxplot(residuals(bdt1),main='Residual Box Plot for beta-carotene')

summary(rdt1)
anova(rdt1)
summary(bdt1)
anova(bdt1)
```

**Polynomial Regression model**
```{r}
rdt2<-lm(retplasma~calories+fat+alcohol+cholesterol+vituse+I(calories^2)+I(fat^2)+I(alcohol^2)+I(cholesterol^2)+
           calories:alcohol+fat:fiber+vituse:betadiet,data=plasma)
bdt2<-lm(betaplasma~betadiet+fat+alcohol+fiber+vituse+I(fiber^2)+I(fat^2)+I(alcohol^2)+I(betadiet^2)
         +vituse:fiber+fiber:betadiet+fat:betadiet+fat:alcohol+calories:alcohol+vituse:alcohol,data=plasma)

# Assumption check
par(mfrow = c(2, 3)) # Set the plots as 1x3 layout

# Residuals VS fitted
plot(x=rdt2$fitted.values,y=residuals(rdt2),main = "Residuals VS Fitted value for retinol",
     xlab = "Fitted value",ylab = "Residuals")
abline(h=0,col='red')

# QQ plot
qqnorm(residuals(rdt2),main='Normal QQ-plot for retinol')
qqline(residuals(rdt2),col='red')

# Residual Box plot
boxplot(residuals(rdt2),main='Residual Box Plot for retinol')

# Residuals VS fitted
plot(x=bdt2$fitted.values,y=residuals(bdt2),main = "Residuals VS Fitted value for beta-carotene",
     xlab = "Fitted value",ylab = "Residuals")
abline(h=0,col='red')

# QQ plot
qqnorm(residuals(bdt2),main='Normal QQ-plot for beta-carotene')
qqline(residuals(bdt2),col='red')

# Residual Box plot
boxplot(residuals(bdt2),main='Residual Box Plot for beta-carotene')

summary(rdt2)
anova(rdt2)
summary(bdt2)
anova(bdt2)
```

**Log-transformed Regression model**
```{r}
# Selection transformation model
anova(rdt,rdt1,rdt2)
anova(bdt,bdt1,bdt2)

rdt3<-lm(log(retplasma)~vituse+calories+fat+fiber+alcohol+cholesterol+retdiet+betadiet
         +calories:alcohol+fat:fiber+vituse:betadiet,data=plasma)
bdt3<-lm(log(betaplasma)~vituse+calories+fat+fiber+alcohol+cholesterol+retdiet+betadiet+
           vituse:fiber+fiber:betadiet+fat:betadiet+fat:alcohol+calories:alcohol+vituse:alcohol,data=plasma)

# Assumption check
par(mfrow = c(2, 3)) # Set the plots as 1x3 layout

# Residuals VS fitted
plot(x=rdt3$fitted.values,y=residuals(rdt3),main = "Residuals VS Fitted value for retinol",
     xlab = "Fitted value",ylab = "Residuals")
abline(h=0,col='red')

# QQ plot
qqnorm(residuals(rdt3),main='Normal QQ-plot for retinol')
qqline(residuals(rdt3),col='red')

# Residual Box plot
boxplot(residuals(rdt3),main='Residual Box Plot for retinol')

# Residuals VS fitted
plot(x=bdt3$fitted.values,y=residuals(bdt3),main = "Residuals VS Fitted value for beta-carotene",
     xlab = "Fitted value",ylab = "Residuals")
abline(h=0,col='red')

# QQ plot
qqnorm(residuals(bdt3),main='Normal QQ-plot for beta-carotene')
qqline(residuals(bdt3),col='red')

# Residual Box plot
boxplot(residuals(bdt3),main='Residual Box Plot for beta-carotene')

summary(rdt3)
anova(rdt3)
summary(bdt3)
anova(bdt3)
```

**Model Selection**
```{r}
# AIC
aic_r<-c(AIC(rdt),AIC(rdt1),AIC(rdt2),AIC(rdt3))
aic_b<-c(AIC(bdt),AIC(bdt1),AIC(bdt2),AIC(bdt3))

#BIC
bic_r<-c(BIC(rdt),BIC(rdt1),BIC(rdt2),BIC(rdt3))
bic_b<-c(BIC(bdt),BIC(bdt1),BIC(bdt2),BIC(bdt3))


comp1<-data.frame(Model = c("Simple Linear", "Interaction Terms", "Polynoimal","Log-Transformed"),
  AIC_ret = aic_r,
  BIC_ret = bic_r,
  AIC_beta=aic_b,
  BIC_beta=bic_b)
comp1

# Outliers concern
library(car)
par(mfrow=c(2,6))
term<-c('vituse','calories','fat','fiber','alcohol','cholesterol','retdiet','betadiet',
         'calories:alcohol','fat:fiber','vituse:betadiet')
for (i in term) {
    leveragePlot(rdt3, term.name = i,id = list(n = 2, col = "red"))
}

par(mfrow=c(2,7))
terms<-c('vituse','calories','fat','fiber','alcohol','cholesterol','retdiet','betadiet',
           'vituse:fiber','fiber:betadiet','fat:betadiet','fat:alcohol','calories:alcohol','vituse:alcohol')
for (i in terms) {
    leveragePlot(bdt3, term.name = i,id = list(n = 2, col = "red"))
}

```